---
title: "Predicting Weight Lifting Exercise Type"
author: "Michael Bausor"
date: "November 22, 2015"
output: html_document
---

## Synopsis

This report describes the process of creating a model in R to predict the type of weight lifting exercise being performed by participants. A random forest model was created using Caret package. A test set consisting of 30% of the data was set aside to assess the out of sample error of the model once training was complete. The model used 52 out of 160 variables to classify the data. The model was found to have an accuracy of 99.4% on the test set.

## Data Processing and Exploratory Analysis

The source of the data was this project:
http://groupware.les.inf.puc-rio.br/har

The training data are available for download here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Once obtained, the data were loaded into R, and partitioned into a training set (70%) and test set (30%) using the following R code:

```{r load}
library(caret)
library(randomForest)
set.seed(1)
data<-read.csv('pml-training.csv')
inTrain <- createDataPartition(y=data$classe, p=0.7, list=FALSE)
training <- data[inTrain,]
testing <- data[-inTrain,]
```

We then reviewed the data, to determine how to proceed. In the dataset there are 160 variables. However, many of these are not suitable or not useful for our analysis. For example, many of the variables were such that almost all of the values were missing. This was true for many of the summary variables, such as avg_roll_arm, stddev_roll_arm, var_roll_arm, etc. as shown in the table below.

```{r tab1}
head(training[,c("avg_roll_arm", "stddev_roll_arm", "var_roll_arm")])
```

Other variables such as the new_window or timestamp are not useful for our analysis, and so can be excluded. We have also chosen to exclude the name of the participants, with the idea that by excluding this information our prediction algorithm may be more likely to generalize to other participants in future applications.

The result of our preliminary analysis is to select 52 of the 160 variables to use for training our model that are the most appropriate for the task. The full list of variables used is:

```{r names, echo=FALSE}
variable_names <- c("roll_belt","pitch_belt","yaw_belt","total_accel_belt","gyros_belt_x","gyros_belt_y","gyros_belt_z","accel_belt_x","accel_belt_y","accel_belt_z","magnet_belt_x","magnet_belt_y","magnet_belt_z","roll_arm","pitch_arm","yaw_arm","total_accel_arm","gyros_arm_x","gyros_arm_y","gyros_arm_z","accel_arm_x","accel_arm_y","accel_arm_z","magnet_arm_x","magnet_arm_y","magnet_arm_z","roll_dumbbell","pitch_dumbbell","yaw_dumbbell","total_accel_dumbbell","gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z","accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z","magnet_dumbbell_x","magnet_dumbbell_y","magnet_dumbbell_z","roll_forearm","pitch_forearm","yaw_forearm","total_accel_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z","accel_forearm_x","accel_forearm_y","accel_forearm_z","magnet_forearm_x","magnet_forearm_y","magnet_forearm_z")
```
```{r tab2}
variable_names
```

## Training and Cross Validation

We will use the Caret package to train our random forest model, with the code below. The details of the model are also shown

```{r train, cache=TRUE}
modelFit <- train(training$classe ~ ., method="rf", data=training[,variable_names])
modelFit
```

Cross validation is performed by the Caret package. We see that several values of the mtry variable have been used. This value represents the number of variables to try at each split in the tree. The most accurate model, selected using cross validation performed by the Caret package, was when mtry=27. The details of the final mode are:

```{r finmod}
modelFit$finalModel
```

## Out of Sample Error

Having trained our model, we can estimate the out of sample error by using it to classify data from the test set that we have set aside, and that represents 30% of our data. We perform this assessment using the r code below. This shows that we have an estimated out of sample accuracy of 99.4%, and therefore an estimated out of sample error of 0.6%

```{r conf}
confusionMatrix(testing$classe, predict(modelFit, testing))
```

## Predictions

Finally, we will use our model to generate predictions for the 20 samples that are available for download here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The R code for the predictions and the results are shown below.

```{r pred}
test_small<-read.csv('pml-testing.csv')
predict(modelFit, test_small)
```